Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-05 resubmission/paper-resubmit.tex
Up render html y pdf

We adopt a multidimensional perspective on meritocracy, distinguishing
between two key dimensions: effort-based and talent-based perceptions.
This distinction is essential, as it captures different pathways through
which individuals justify inequality. Effort-based meritocracy
emphasizes hard work and perseverance as the basis for social rewards,
aligning closely with cultural narratives of personal responsibility. A
talent-based meritocracy, by contrast, emphasizes intelligence and
innate abilities, which are often perceived as less malleable and more
unequally distributed. Both dimensions have been shown to correlate with
acceptance of inequality, but they may carry distinct implications for
how inequality is justified in specific domains
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
The relevance of this distinction is supported by recent studies, which
show that individuals respond differently to these dimensions. For
instance, perceptions that effort is rewarded in society are more
strongly associated with positive evaluations of fairness and acceptance
of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}). This may be because effort is seen as a controllable and morally
virtuous trait, whereas talent is often perceived as a natural
advantage. Consequently, effort-based meritocracy is likely more potent
in legitimizing inequality, particularly in neoliberal contexts.

@@

Second, higher perceived meritocracy is expected to be positively
associated with market justice preferences. Individuals who believe that
effort and talent primarily determine success are more likely to justify
unequal outcomes and endorse the notion that markets allocate resources
fairly according to individual merit
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
This aligns with the idea that meritocratic perceptions serve as a moral
framework that legitimizes market-based inequalities, as individuals
perceive the system as just when they believe that rewards are based on
individual merit. This is particularly relevant in contexts where
neoliberal ideologies dominate, as they often emphasize individual
responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over
time, particularly in light of broader social and political
transformations. We are especially interested in whether changes in
individuals' perceptions of inequality and meritocracy are associated
with shifts in their market justice attitudes across different time
points. The primary focus of this study is thus to scrutinize how
individual-level changes in perceptions relate to changes in preferences
over time. While we acknowledge the relevance of contextual events ---
such as the 2019 political outburst or the subsequent constitutional
process --- for public opinion and political orientation
(\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
2025}), identifying causal effects of such events (e.g., mass protest
exposure) on individual attitudes lies outside the scope of this
research. Instead, we focus on how intra-individual trajectories in
perceived inequality and merit shape evolving attitudes toward the
commodification of social services.

@@

The first hypothesis suggested that perceptions of economic inequality
were strongly linked to preferences for market justice. The findings
demonstrated that individuals perceiving higher income disparities
showed less support for the idea that the access to core social services
depends on individual income, aligning with theories emphasizing that
awareness of inequality could fuel a critical view of such policies
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
reflected a moral economy where perceptions of systemic unfairness
threatened the legitimacy of existing distributions, prompting calls for
greater equity. Similarly, the negative within-effects observed over
time (\(H1_{b}\)) implied that as perceptions of inequality became more
salient, some individuals experienced a decline in support for market
justice, possibly due to increasing distrust in market mechanisms or
disillusionment with the fairness of the system.

Regarding the second hypothesis, the results reinforced that
meritocratic perceptions ---particularly those emphasizing individual
effort--- were associated with stronger support for market-based
distribution systems. Individuals who viewed success as primarily earned
through effort were more inclined to justify existing inequalities,
seeing them as outcomes of individual merit rather than systemic
injustice. This aligned with research indicating that meritocratic
narratives served as moral justifications that reinforced societal
stratification (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
functioned symbolically to legitimize structural disparities, reducing
support for redistributive measures by framing inequality as fair and
deserved. The findings suggested that these perceptions bolstered the
social hierarchy by fostering acceptance of inequality as a reflection
of individual virtue rather than structural failure, which was
especially pertinent in neoliberal contexts like Chile's, where market
logic heavily influenced social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
Interestingly, intraindividual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While an increase in the
perception that rewards are based on individual effort is associated
with stronger preferences for market fairness, an increase in the belief
that rewards come from talent is linked to lower support for such
principles. One possible explanation is that effort is generally
perceived as a controllable and malleable trait, which aligns with the
normative foundations of market fairness. Conversely, talent may be
viewed as an innate and less alterable attribute, making talent-based
reward allocation seem less fair or legitimate. Moreover, greater
exposure to real-life situations where outcomes are strongly determined
by inherent characteristics, rather than effort or hard work, may lead
people to question the fairness of market-based reward systems.

The third hypothesis concerned the moderating role of perceptions of
inequality on the relationship between meritocracy and market justice
attitudes. The analysis indicated that the positive association between
meritocratic perceptions and support for market-based distribution
tended to strengthen when perceptions of economic inequality were
particularly high. While this moderation is not significant for
effort-related effects, it is significant for talent in the within- and
between-individual effects. This appears as contrary to our initial
hypothesis, as it would mean that effect of meritocratic perceptions
becomes stronger as perception of inequality increases. Nevertheless, we
need to be cautious about this interpretation, as the main effect of
inequality perception on market justice is negative. Therefore, as
perceived inequality become less negative (i.e.~more positive), or
closer to zero, then the meritocratic perception has an stronger
positive relationship on market justice preferences. This could mean
that meritocratic perceptions play an stronger role in justification of
market justice inequalities for those who are less capable or less prone
to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social
movements---particularly those that emerged in 2019---on the
relationship between meritocratic beliefs, perceptions of inequality,
and attitudes toward market justice. The main effect of time suggests
that, on average, support for market justice increased after 2019.
However, the results do not reveal any significant interactions between
distributive mechanisms and market justice attitudes, indicating that
the 2019 social mobilizations did not substantially alter the underlying
associations between these variables. This may suggest a certain
stability or resilience in the normative frameworks that shape
individuals' evaluations of justice, even in the context of major
collective political events. Alternatively, it is possible that while
the protests challenged institutional legitimacy, they did not
fundamentally shift individuals' views regarding how rewards should be
allocated in society.

>>>>

We adopt a multidimensional perspective on meritocracy, distinguishing
between two key dimensions: effort-based and talent-based perceptions.
This distinction is essential, as it captures different pathways through
which individuals justify inequality. Effort-based meritocracy
emphasizes hard work and perseverance as the basis for social rewards,
aligning closely with cultural narratives of personal responsibility. A
talent-based meritocracy, by contrast, emphasizes intelligence and
innate abilities, which are often perceived as less malleable and more
unequally distributed. Both dimensions have been shown to correlate with
acceptance of inequality, but they may carry distinct implications for
how inequality is justified in specific domains
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
The relevance of this distinction is supported by recent studies, which
show that individuals respond differently to these dimensions. For
instance, perceptions that effort is rewarded in society are more
strongly associated with positive evaluations of fairness and acceptance
of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}).

@@

Second, higher perceived meritocracy is expected to be positively
associated with market justice preferences. Individuals who believe that
effort and talent primarily determine success would be more likely to
justify unequal outcomes and endorse the notion that markets allocate
resources fairly according to individual merit
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
This aligns with the idea that meritocratic perceptions serve as a moral
framework that legitimizes market-based inequalities, as individuals
perceive the system as just when they believe that rewards are based on
individual merit. This is particularly relevant in contexts where
neoliberal ideologies dominate, as they often emphasize individual
responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over
time, particularly in light of broader social and political
transformations. We are especially interested in whether changes in
individuals' perceptions of inequality and meritocracy are associated
with shifts in their market justice attitudes across different time
points. While we acknowledge the relevance of contextual events --- such
as the 2019 political outburst or the subsequent constitutional process
--- for public opinion and political orientation
(\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
2025}), identifying causal effects of such events (e.g., mass protest
exposure) on individual attitudes lies outside the scope of this
research. Instead, we focus on how intra-individual trajectories in
perceived inequality and merit shape evolving attitudes toward the
commodification of social services.

@@

The first set of hypothesis suggested that perceptions of economic
inequality were strongly linked to preferences for market justice. The
findings demonstrated that individuals perceiving higher income
disparities showed less support for the idea that the access to core
social services depends on individual income, aligning with theories
emphasizing that awareness of inequality could fuel a critical view of
such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
reflects a moral economy in which perceptions of systemic unfairness
undermine the legitimacy of existing distributions and fuel demands for
greater equity. In particular, the negative within-individual effect
observed over time (\(H1_{b}\)) suggests that as inequality becomes more
salient for some individuals, their support for market justice
declines---potentially due to rising distrust in market mechanisms or
growing disillusionment with the fairness of the system. However,
although the effect of perceived inequality at the between-individual
level is also negative, it is not statistically significant. Therefore,
we cannot conclude that individuals who, on average, perceive higher
inequality are systematically less supportive of market justice. As a
result, hypothesis \(H1_{a}\) does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that
meritocratic perceptions---particularly those emphasizing individual
effort---were associated with stronger support for market-based
distribution systems. Individuals who believed that success is primarily
achieved through personal effort were more likely to justify unequal
access to core social services based on income, interpreting such
disparities as outcomes of individual merit rather than systemic
injustice (\(H2_{a}\)). This finding aligns with previous research
showing that meritocratic narratives serve as moral justifications that
legitimize social stratification
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
operate symbolically to reinforce structural inequalities by reducing
support for redistributive policies, framing inequality as both fair and
deserved. Consistent with prior work by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
the findings suggest that such meritocratic perceptions uphold existing
hierarchies by promoting the acceptance of inequality as a reflection of
individual virtue rather than structural failure. This mechanism is
particularly salient in neoliberal contexts like Chile, where market
logics heavily shape social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

Interestingly, intra-individual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
that rewards are based on individual effort are associated with stronger
preferences for market justice, increases in the belief that rewards
derive from talent are linked to weaker support for such principles. One
possible explanation is that effort is generally viewed as a
controllable and malleable trait, aligning with the normative
foundations of market fairness. In contrast, talent tends to be
perceived as innate and less subject to personal control, rendering
talent-based inequality less legitimate. Furthermore, increased exposure
to real-world scenarios in which outcomes are clearly shaped by inherent
traits rather than hard work may lead individuals to question the
fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions
of inequality in the relationship between meritocratic perceptions and
preferences for market justice. The analysis revealed that the positive
association between meritocratic perceptions and support for
market-based distribution systems tends to be stronger when perceived
economic inequality is high---contradicting our initial expectations
(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
statistically significant for effort-related meritocratic perceptions,
it is significant for talent-related perceptions at both the within- and
between-individual levels. This finding runs counter to our hypotheses,
as it suggests that the effect of meritocratic
perceptions---specifically, talent-based meritocracy---intensifies as
perceptions of inequality increase. However, this interpretation
requires caution. The main effect of perceived inequality on support for
market justice is negative. Thus, as perceived inequality becomes less
negative (i.e., closer to zero), the positive relationship between
talent-based meritocratic perceptions and market justice preferences
becomes stronger. This suggests a more nuanced dynamic: meritocratic
perceptions, particularly those emphasizing talent, may serve as a
stronger justificatory mechanism for market-based inequalities among
individuals who are less inclined to perceive inequality as a problem.
In other words, among those who do not strongly perceive systemic
disparities, meritocratic narratives may play a more influential role in
legitimizing unequal outcomes.

Finally, the fourth hypothesis examined the potential impact of recent
social mobilizations---particularly those that erupted in 2019---on the
relationship between meritocratic perceptions, perceptions of
inequality, and preferences for market justice. The main effect of time
indicates that, on average, support for market justice increased
following the 2019 protests. However, the analysis did not detect any
significant interactions between distributive beliefs and support for
market justice, suggesting that the social mobilizations did not
substantially modify the underlying associations between these
variables. As a result, hypotheses \(H4_{a}\) and \(H4_{b}\) are not
supported by the data. This may point to a certain stability---or even
resilience---in the normative frameworks that guide individuals'
evaluations of distributive justice, despite the occurrence of major
collective political events. Alternatively, while the protests may have
eroded trust in institutional arrangements or governance, they may not
have fundamentally altered individuals' beliefs about how rewards and
resources should be allocated in society.
